---
title: "data wrangling intro"
author: "Julia Parish"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)

```

# Warm-up

Look at `us_rent_income` (built-in to the `tidyr` package). Make a tidy version. Then from your tidy version, get it back to the original structure. 

```{r}
us_rent_wide <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable,
              values_from = estimate)
```

## Practice some tidyverse functions for data wrangling
- `filter`
- `select`
- `mutate()`
- `case_when()`
- `group_by() %>% summarize`
- piping things together in sequence

### dplyr::filter() to keep or exclude rows based on our conditions

1. Exact match (single cond't)

Only keep observations where the island is Biscoe
```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
unique(ex_1$island)

```

2. "and" statement within filter

Only keep observations if the penguin species is chinstrap AND the year is 2008
```{r}
# 1st option
ex_2 <- penguins %>% 
  filter(species == "Chinstrap", year == 2008)

# 2nd option
ex_2 <- penguins %>% 
  filter(species == "Chinstrap" & year == 2008)

# 3rd option
ex_2 <- penguins %>% 
  filter(species == "Chinstrap") %>% 
  filter(year == 2008)
```
# 3. Excliude Gentoos

keep obersvations if the penguin species is NOT "Gentoo"

```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
  
```
# Excercise:
- Create a subset from `penguins` that only contains observations for the Adelie penguins with a flipper length greater than or equal to 185mm. 

```{r}

adelie_subset <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)

unique(adelie_subset)

```
# 4. The %in% operator
- use the %in% operator to look for *any* mateches within a vector, and return TRUE if the value does exist for our row.
- We want to keep observations where the island is Dream or Torgersen. 

```{r}
ex_4 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))
```
# What would actually be happening if in the code above, I had used `==` instead of `%in%`. 


for just Biscoe - not Dream or Torgersen
```{r}
ex_4 <- penguins %>% 
  filter(!island %in% c("Dream", "Torgersen"))
```

### Select is to select of exclude columns
# 5. Keep a single column by name
Only want to keep the species column

```{r}
ex_5 <- penguins %>% select(species) %>% 
  pull(species)

# write this without the pipe operator

ex_5a <- select(.data = penguins, species)

```

# 6. Keep multiple columns by name
- Create a subset that contains species, bill_length_mm thru flipper_length_mm, and year

```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)

```

# 7. Choose or excludes based on position
- exclude column 4 in penguins. 
```{r}
ex_7 <- penguins %>% 
  select(-4)
```

# 8. A piped sequence 
- write a piped sequence where first only keep columns for species, body_mass_g, and island, and second: filter to only includde body_mass_g is greater than 3600

```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)
```

### Mutate() to add or update existing columns, while keeping the ones already there
# 9. Add a column that has body mass converted to kg
- meant to add flipper_length_mm converted to meters, so add another mutate!
```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) %>% 
  mutate(flipper_length_mm = flipper_length_mm / 1000)

```

# 10. Mutate to update the existing column - create new version of island using string replace looking for pattern sen and the replacement is hooray!
```{r}
ex_10 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", replacement = "HOORAY"))

# if you don't want to overwrite a column, but create a new one with updated data
ex_10 <- penguins %>% 
  mutate(island_HOORAY = str_replace(island, pattern = "sen", replacement = "HOORAY"))

```

11. Update the class of the bill_depth column to character
is.numeric is asking a question, as.numeric changes

```{r}
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))

class(ex_11$bill_depth_mm)
```
### case_when(): your favorite if-else
# 12. Create a new column that contains "small" if penguin flipper length is less than 185 mm, "medium" otherwise
```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
  ))
  
```
# Task - starting from penguins, create a new column "bird_bill_bin" that contains:
- If the bill length < 38mm: "short bill"
- If the bill length >= 38mm & < 43 mm: "medium bill"
- If the bill lengeth >= 43mm: "long bill"
```{r}
ex_12a <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill",
    bill_length_mm >= 43 ~ "long bill"
  ))
```

### group_by() %>% summarize() to find summary values by group, and return them in a nice summary table
- what is the mean value for bill depth for all species
```{r}
mean(penguins$bill_depth_mm, na.rm = TRUE)

```
# group_by() to say there are groups within the species column
- then add by island
```{r}
ex_13 <- penguins %>% 
  group_by(species, island) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))
ex_13

# mutate does something very different - creates a mean flipper length column for the mean flipper length for each species on each row. Does not create a summary table, rather keeps orginal data frame.
penguins %>% 
  group_by(species) %>% 
  mutate(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE))


```

# Practice task
In a single pipe sequence:
- only keep penguins that have a body mass > 3500 g
- only keep columns for species, flipper_length_mm, and sex
- create a summary table that contains the mean, sd, max and min value of the flipper length by species and sex (as the grouping variables)

```{r}
practice_3 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE),
            max_flip = max(flipper_length_mm, na.rm = TRUE),
            min_flip = min(flipper_length_mm, na.rm = TRUE)) %>% 
  ungroup() %>% 
  drop_na(sex)
  
min(practice_3$body_mass_g)  
```
